---
title: "Brim AL Portfolio Investor Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme:
      version: 4
      bootswatch: flatly
      primary: "#1e3a8a"
    css: dashboard_styles.css
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(scales)
library(DT)

# Load processed dashboard data
dashboard_data <- readRDS("../data/processed/dashboard_data.rds")

kpi <- dashboard_data$kpi
t12 <- dashboard_data$t12
valuation_long <- dashboard_data$valuation_long
loan_long <- dashboard_data$loan_long

# Color palette
colors <- list(
  primary = "#1e3a8a",
  success = "#059669",
  warning = "#d97706",
  danger = "#dc2626",
  fund_i = "#3b82f6",
  fund_ii = "#10b981",
  egi = "#6366f1",
  fixed = "#f59e0b",
  variable = "#ef4444",
  noi = "#10b981"
)
```

Executive Summary
=====================================

Row {data-height=120}
-----------------------------------------------------------------------

### Total Capital Invested

```{r}
valueBox(
  value = dollar(kpi$total_icc, accuracy = 1),
  caption = "Total I.C.C.",
  icon = "fa-dollar-sign",
  color = "primary"
)
```

### Portfolio Units

```{r}
valueBox(
  value = kpi$unit_count,
  caption = "Single Family Homes",
  icon = "fa-home",
  color = "info"
)
```

### 2026 Potential NOI

```{r}
valueBox(
  value = dollar(kpi$potential_noi_2026, accuracy = 1),
  caption = "Projected Annual NOI",
  icon = "fa-chart-line",
  color = "success"
)
```

### Potential Cap Rate

```{r}
valueBox(
  value = percent(kpi$potential_cap_rate, accuracy = 0.1),
  caption = "At Stabilization",
  icon = "fa-percentage",
  color = "warning"
)
```

### NOI as of Oct 2025

```{r}
current_noi <- t12$noi[nrow(t12)]
valueBox(
  value = dollar(current_noi, accuracy = 1),
  caption = "NOI as of Oct 2025",
  icon = "fa-money-bill-wave",
  color = "info"
)
```

### Cap Rate as of Oct 2025

```{r}
current_cap <- t12$cap_rate[nrow(t12)]
valueBox(
  value = percent(current_cap, accuracy = 0.1),
  caption = "Cap Rate as of Oct 2025",
  icon = "fa-chart-pie",
  color = "primary"
)
```

Row {data-height=380}
-----------------------------------------------------------------------

### Fund Capital Allocation

```{r}
fund_data <- tibble(
  Fund = c("Fund I", "Fund II"),
  Capital = c(kpi$fund_i_icc, kpi$fund_ii_icc),
  Percent = c(kpi$fund_i_icc, kpi$fund_ii_icc) / kpi$total_icc
)

plot_ly(fund_data, 
        labels = ~Fund, 
        values = ~Capital, 
        type = 'pie',
        hole = 0.5,
        textinfo = 'label+percent',
        textposition = 'outside',
        marker = list(colors = c(colors$fund_i, colors$fund_ii)),
        hovertemplate = "%{label}<br>%{value:$,.0f}<br>%{percent}<extra></extra>") |>
  layout(
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.1),
    annotations = list(
      text = paste0("<b>", dollar(kpi$total_icc, scale = 1e-6, suffix = "M"), "</b><br>Total"),
      showarrow = FALSE,
      font = list(size = 16)
    )
  )
```

### T-12 Financial Performance Trend

```{r}
t12_long <- t12 |>
  select(month, egi, fixed_expense, variable_expense, noi) |>
  pivot_longer(-month, names_to = "metric", values_to = "value") |>
  mutate(
    metric = case_when(
      metric == "egi" ~ "Effective Gross Income",
      metric == "fixed_expense" ~ "Fixed Expenses",
      metric == "variable_expense" ~ "Variable Expenses",
      metric == "noi" ~ "Net Operating Income"
    ),
    metric = factor(metric, levels = c("Effective Gross Income", "Fixed Expenses", 
                                        "Variable Expenses", "Net Operating Income"))
  )

color_map <- c(
  "Effective Gross Income" = colors$egi,
  "Fixed Expenses" = colors$fixed,
  "Variable Expenses" = colors$variable,
  "Net Operating Income" = colors$noi
)

plot_ly(t12_long, x = ~month, y = ~value, color = ~metric, 
        colors = color_map, type = 'scatter', mode = 'lines+markers',
        hovertemplate = "%{x}<br>%{y:$,.0f}<extra>%{fullData.name}</extra>") |>
  layout(
    xaxis = list(title = ""),
    yaxis = list(title = "", tickformat = "$,.0f"),
    legend = list(orientation = "h", y = -0.15),
    hovermode = "x unified"
  )
```

Row {data-height=380}
-----------------------------------------------------------------------

### EGI Breakdown: Expenses vs NOI

```{r}
# Stacked bar with clear visual separation between expenses (red tones) and NOI (green)
expense_data <- t12 |>
  mutate(
    total_expenses = fixed_expense + variable_expense,
    fixed_pct = fixed_expense / egi * 100,
    variable_pct = variable_expense / egi * 100,
    noi_pct = noi / egi * 100
  )

plot_ly(expense_data, x = ~month) |>
  # Expenses section (warm colors, stacked together)
  add_trace(y = ~fixed_expense, name = "Fixed Expenses", 
            type = 'bar',
            marker = list(color = "#fbbf24", line = list(color = "#f59e0b", width = 1)),
            text = ~paste0(round(fixed_pct, 0), "%"),
            textposition = "inside",
            hovertemplate = "Fixed Expenses<br>%{y:$,.0f} (%{text})<extra></extra>") |>
  add_trace(y = ~variable_expense, name = "Variable Expenses", 
            type = 'bar',
            marker = list(color = "#f87171", line = list(color = "#ef4444", width = 1)),
            text = ~paste0(round(variable_pct, 0), "%"),
            textposition = "inside",
            hovertemplate = "Variable Expenses<br>%{y:$,.0f} (%{text})<extra></extra>") |>
  # NOI section (green with distinct styling)
  add_trace(y = ~noi, name = "Net Operating Income", 
            type = 'bar',
            marker = list(
              color = "#34d399",
              line = list(color = "#059669", width = 2),
              pattern = list(shape = "/", solidity = 0.1)
            ),
            text = ~paste0(round(noi_pct, 0), "%"),
            textposition = "inside",
            textfont = list(color = "#065f46", size = 12, weight = "bold"),
            hovertemplate = "NOI (Profit)<br>%{y:$,.0f} (%{text})<extra></extra>") |>
  layout(
    barmode = 'stack',
    xaxis = list(title = ""),
    yaxis = list(title = "Effective Gross Income", tickformat = "$,.0f"),
    legend = list(orientation = "h", y = -0.15, 
                  traceorder = "normal"),
    hovermode = "x unified",
    annotations = list(
      list(x = 0.5, y = 1.08, xref = "paper", yref = "paper",
           text = "<b>EXPENSES</b> (Fixed + Variable) vs <b style='color:#059669'>NOI</b> (Profit)",
           showarrow = FALSE, font = list(size = 11))
    )
  )
```

### Cap Rate Trend

```{r}
plot_ly(t12, x = ~month, y = ~cap_rate, type = 'scatter', mode = 'lines+markers',
        line = list(color = colors$primary, width = 3),
        marker = list(size = 10, color = colors$primary),
        hovertemplate = "%{x}<br>Cap Rate: %{y:.2%}<extra></extra>") |>
  add_trace(y = rep(kpi$potential_cap_rate, nrow(t12)), 
            name = "Target (9.16%)",
            line = list(color = colors$success, width = 2, dash = "dash"),
            marker = list(size = 0),
            hovertemplate = "Target: %{y:.2%}<extra></extra>") |>
  layout(
    xaxis = list(title = ""),
    yaxis = list(title = "Capitalization Rate", tickformat = ".1%", 
                 range = c(0, max(kpi$potential_cap_rate * 1.1, 0.1))),
    legend = list(orientation = "h", y = -0.15),
    annotations = list(
      list(x = "Oct", y = kpi$potential_cap_rate, 
           text = "Stabilized Target", showarrow = FALSE,
           xanchor = "left", yanchor = "bottom",
           font = list(color = colors$success, size = 10))
    )
  )
```


Scenario Analysis
=====================================

Row {data-height=500}
-----------------------------------------------------------------------

### Portfolio Valuation: Current vs Target Position

```{r}
# Current and target positions
current_noi <- t12$noi[nrow(t12)]  # Oct 2025: ~$134K
current_cap <- t12$cap_rate[nrow(t12)] * 100  # ~3.6%
current_value <- current_noi / (current_cap / 100)

target_noi <- kpi$potential_noi_2026  # ~$260K
target_cap <- kpi$potential_cap_rate * 100  # ~9.16%
target_value <- target_noi / (target_cap / 100)

# Focus on relevant range around current and target
noi_range <- seq(100000, 300000, by = 20000)
cap_range <- seq(3, 10, by = 0.5)

# Build grid with progress score for coloring
# Color logic: Higher NOI + Higher Cap Rate = GREEN (stabilized/good)
val_grid <- expand_grid(noi = noi_range, cap_rate = cap_range) |>
  mutate(
    value = noi / (cap_rate / 100),
    # Progress score: 0 at bottom-left, 1 at top-right
    noi_norm = (noi - min(noi_range)) / (max(noi_range) - min(noi_range)),
    cap_norm = (cap_rate - min(cap_range)) / (max(cap_range) - min(cap_range)),
    progress = (noi_norm + cap_norm) / 2
  )

# Create heatmap using plot_ly with data frame (more reliable)
plot_ly(val_grid, x = ~noi, y = ~cap_rate, z = ~progress, 
        type = "heatmap",
        colors = colorRamp(c("#dc2626", "#f97316", "#fbbf24", "#84cc16", "#22c55e")),
        text = ~paste0("Value: ", scales::dollar(value)),
        hovertemplate = "NOI: %{x:$,.0f}<br>Cap Rate: %{y:.1f}%<br>%{text}<extra></extra>",
        colorbar = list(title = "Progress", tickvals = c(0, 0.5, 1), 
                        ticktext = c("Start", "Mid", "Target"), len = 0.8),
        showscale = TRUE) |>
  # Current position marker
  add_trace(
    x = current_noi, y = current_cap, type = "scatter", mode = "markers+text",
    marker = list(size = 18, color = "#2563eb", line = list(color = "white", width = 3)),
    text = "CURRENT", textposition = "top center",
    textfont = list(color = "#1e40af", size = 11, family = "Arial Black"),
    hovertemplate = paste0("<b>CURRENT</b><br>NOI: ", dollar(current_noi), 
                           "<br>Cap: ", round(current_cap, 1), "%<br>Value: ", 
                           dollar(current_value), "<extra></extra>"),
    name = "Current (Oct 2025)", inherit = FALSE
  ) |>
  # Target position marker
  add_trace(
    x = target_noi, y = target_cap, type = "scatter", mode = "markers+text",
    marker = list(size = 18, color = "#16a34a", symbol = "diamond",
                  line = list(color = "white", width = 3)),
    text = "TARGET", textposition = "bottom center",
    textfont = list(color = "#166534", size = 11, family = "Arial Black"),
    hovertemplate = paste0("<b>TARGET</b><br>NOI: ", dollar(target_noi), 
                           "<br>Cap: ", round(target_cap, 1), "%<br>Value: ", 
                           dollar(target_value), "<extra></extra>"),
    name = "Target (2026)", inherit = FALSE
  ) |>
  layout(
    xaxis = list(title = "Net Operating Income (NOI)", tickformat = "$,.0f"),
    yaxis = list(title = "Capitalization Rate (%)", autorange = "reversed"),
    legend = list(orientation = "h", y = -0.12, x = 0.5, xanchor = "center"),
    annotations = list(
      list(x = target_noi, y = target_cap, ax = current_noi, ay = current_cap,
           xref = "x", yref = "y", axref = "x", ayref = "y",
           showarrow = TRUE, arrowhead = 2, arrowsize = 1.5, arrowwidth = 2, arrowcolor = "#7c3aed"),
      list(x = 0.5, y = 1.06, xref = "paper", yref = "paper",
           text = "<b>Higher NOI + Higher Cap Rate = GREEN (stabilized)</b>",
           showarrow = FALSE, font = list(size = 11, color = "#374151"))
    )
  )
```

### Loan Affordability by Interest Rate

```{r}
# Focus on practical interest rate range (5-9%) and debt service capacity
ds_range <- seq(80000, 200000, by = 15000)
rate_range <- seq(5, 9, by = 0.5)

# Build grid with affordability score for coloring
# Color logic: Higher DS + Lower Rate = GREEN (more borrowing power)
loan_grid <- expand_grid(ds = ds_range, rate = rate_range) |>
  mutate(
    max_loan = ds / (rate / 100),
    # Affordability score: high DS + low rate = good (top-right in visual)
    ds_norm = (ds - min(ds_range)) / (max(ds_range) - min(ds_range)),
    rate_norm = (max(rate_range) - rate) / (max(rate_range) - min(rate_range)),  # Inverted
    affordability = (ds_norm + rate_norm) / 2
  )

# Current debt service capacity
current_ds <- current_noi * 0.8  # 80% of NOI for debt service (1.25x DSCR)

# Create heatmap using plot_ly with data frame
plot_ly(loan_grid, x = ~ds, y = ~rate, z = ~affordability, 
        type = "heatmap",
        colors = colorRamp(c("#dc2626", "#f97316", "#fbbf24", "#84cc16", "#22c55e")),
        text = ~paste0("Max Loan: ", scales::dollar(max_loan)),
        hovertemplate = "Debt Service: %{x:$,.0f}/yr<br>Rate: %{y:.1f}%<br>%{text}<extra></extra>",
        colorbar = list(title = "Borrowing<br>Power", tickvals = c(0, 0.5, 1), 
                        ticktext = c("Low", "Mid", "High"), len = 0.8),
        showscale = TRUE) |>
  # Mark current DS capacity line
  add_trace(
    x = rep(current_ds, length(rate_range)), y = rate_range,
    type = "scatter", mode = "lines",
    line = list(color = "#2563eb", width = 3, dash = "dash"),
    name = paste0("Current DS (~", dollar(current_ds, accuracy = 1), ")"),
    hovertemplate = "Current DS Capacity<br>%{x:$,.0f}<extra></extra>",
    inherit = FALSE
  ) |>
  layout(
    xaxis = list(title = "Annual Debt Service Capacity", tickformat = "$,.0f"),
    yaxis = list(title = "Interest Rate (%)", autorange = "reversed"),
    legend = list(orientation = "h", y = -0.12),
    annotations = list(
      list(x = 0.5, y = 1.06, xref = "paper", yref = "paper",
           text = "<b>Higher DS + Lower Rate = GREEN (more borrowing power)</b>",
           showarrow = FALSE, font = list(size = 11, color = "#374151"))
    )
  )
```

Row {data-height=380}
-----------------------------------------------------------------------

### Key Metrics Summary

```{r}
metrics_table <- tibble(
  Metric = c(
    "Total Capital Invested (I.C.C.)",
    "Fund I Contribution",
    "Fund II Contribution",
    "Number of Units",
    "Potential Monthly Rent",
    "Potential Annual Rent",
    "2026 Projected NOI",
    "Stabilized Cap Rate",
    "Current T-12 NOI (Oct 2025)",
    "Current Cap Rate"
  ),
  Value = c(
    dollar(kpi$total_icc),
    dollar(kpi$fund_i_icc),
    dollar(kpi$fund_ii_icc),
    as.character(kpi$unit_count),
    dollar(kpi$potential_monthly_rent),
    dollar(kpi$potential_annual_rent),
    dollar(kpi$potential_noi_2026),
    percent(kpi$potential_cap_rate, accuracy = 0.01),
    dollar(t12$noi[nrow(t12)]),
    percent(t12$cap_rate[nrow(t12)], accuracy = 0.01)
  ),
  Category = c(
    rep("Capital", 3),
    rep("Portfolio", 4),
    rep("Returns", 3)
  )
)

datatable(
  metrics_table,
  options = list(
    pageLength = 15,
    dom = 't',
    ordering = FALSE
  ),
  rownames = FALSE,
  class = 'cell-border stripe'
) |>
  formatStyle(
    'Category',
    backgroundColor = styleEqual(
      c('Capital', 'Portfolio', 'Returns'),
      c('#eff6ff', '#f0fdf4', '#fef3c7')
    )
  )
```

### T-12 Monthly Data

```{r}
t12_display <- t12 |>
  mutate(
    `Effective Gross Income` = dollar(egi),
    `Fixed Expenses` = dollar(fixed_expense),
    `Variable Expenses` = dollar(variable_expense),
    `Net Operating Income` = dollar(noi),
    `Cap Rate` = percent(cap_rate, accuracy = 0.01),
    `Month` = as.character(month)
  ) |>
  select(Month, `Effective Gross Income`, `Fixed Expenses`, 
         `Variable Expenses`, `Net Operating Income`, `Cap Rate`)

datatable(
  t12_display,
  options = list(
    pageLength = 6,
    dom = 't',
    ordering = FALSE
  ),
  rownames = FALSE,
  class = 'cell-border stripe'
)
```


Calculators
=====================================

Row {data-height=500}
-----------------------------------------------------------------------

### Portfolio Valuation Calculator

```{r}
# Reference values
portfolio_value <- kpi$total_icc  # Fixed portfolio value ~$2.835M
current_noi_ref <- t12$noi[nrow(t12)]  # ~$134K
target_noi_ref <- kpi$potential_noi_2026  # ~$260K
current_cap_ref <- current_noi_ref / portfolio_value * 100  # Calculated from NOI
target_cap_ref <- target_noi_ref / portfolio_value * 100  # Calculated from NOI

fillCol(height = "100%", flex = c(NA, 1),

  div(style = "padding: 20px; background: #f8fafc; border-radius: 8px; margin-bottom: 15px;",
    h4("Adjust NOI to see Cap Rate (Portfolio Value is fixed)", style = "margin-top: 0; color: #1e3a8a;"),
    fluidRow(
      column(8, offset = 2,
        sliderInput("noi_input", "Net Operating Income (NOI)",
                    min = 50000, max = 350000, value = round(current_noi_ref, -3),
                    step = 5000, pre = "$", sep = ",",
                    width = "100%")
      )
    ),
    fluidRow(
      column(12, align = "center",
        p(style = "margin: 5px 0; color: #6b7280; font-size: 12px;",
          paste0("Portfolio Value (I.C.C.): ", dollar(portfolio_value), " | ",
                 "Current NOI: ", dollar(current_noi_ref), " → ", round(current_cap_ref, 2), "% | ",
                 "Target NOI: ", dollar(target_noi_ref), " → ", round(target_cap_ref, 2), "%"))
      )
    )
  ),
  renderUI({
    noi <- input$noi_input
    
    # Cap Rate auto-calculated from NOI and fixed Portfolio Value
    cap_rate <- noi / portfolio_value * 100
    
    # Color logic: Higher NOI = Higher Cap Rate = GREEN (good/stabilized)
    progress <- (noi - current_noi_ref) / (target_noi_ref - current_noi_ref)
    progress <- max(0, min(1, progress))  # Clamp 0-1
    
    # Color gradient: Red (0) -> Yellow (0.5) -> Green (1)
    if (progress < 0.5) {
      r <- 220
      g <- round(180 * (progress * 2))
      b <- 50
    } else {
      r <- round(220 - 180 * ((progress - 0.5) * 2))
      g <- 180
      b <- 50
    }
    bg_color <- sprintf("rgb(%d, %d, %d)", r, g, b)
    text_color <- if (progress > 0.3) "#1a1a1a" else "#ffffff"
    
    # Status label
    status <- case_when(
      progress >= 0.9 ~ "At Target",
      progress >= 0.7 ~ "Near Target",
      progress >= 0.4 ~ "In Progress",
      progress >= 0.2 ~ "Below Target",
      TRUE ~ "Starting Point"
    )
    
    div(style = "text-align: center; padding: 30px;",
      # Two side-by-side result boxes
      div(style = "display: flex; justify-content: center; gap: 30px; flex-wrap: wrap;",
        # Cap Rate box (primary result)
        div(style = paste0(
          "padding: 30px 50px; border-radius: 16px; ",
          "background: ", bg_color, "; ",
          "box-shadow: 0 10px 40px rgba(0,0,0,0.15);"
        ),
          p(style = paste0("margin: 0 0 8px 0; font-size: 14px; color: ", text_color, "; opacity: 0.8;"),
            "Cap Rate"),
          h1(style = paste0("margin: 0; font-size: 42px; font-weight: bold; color: ", text_color, ";"),
             percent(cap_rate/100, accuracy = 0.01)),
          p(style = paste0("margin: 8px 0 0 0; font-size: 13px; color: ", text_color, "; opacity: 0.9;"),
            status)
        ),
        # Portfolio Value box (fixed reference)
        div(style = "padding: 30px 50px; border-radius: 16px; background: #e2e8f0; box-shadow: 0 4px 15px rgba(0,0,0,0.08);",
          p(style = "margin: 0 0 8px 0; font-size: 14px; color: #64748b;",
            "Portfolio Value (Fixed)"),
          h1(style = "margin: 0; font-size: 42px; font-weight: bold; color: #1e3a8a;",
             dollar(portfolio_value)),
          p(style = "margin: 8px 0 0 0; font-size: 13px; color: #64748b;",
            "Total I.C.C.")
        )
      ),
      div(style = "margin-top: 25px; color: #6b7280;",
        p(style = "margin: 5px 0;", 
          strong("Formula: "), "Cap Rate = NOI / Portfolio Value"),
        p(style = "margin: 5px 0; font-size: 13px;",
          HTML("Higher NOI → Higher Cap Rate = <span style='color: #16a34a; font-weight: bold;'>GREEN</span> (stabilized)"))
      )
    )
  })
)
```

### Loan Affordability Calculator

```{r}
fillCol(height = "100%", flex = c(NA, 1),
  div(style = "padding: 20px; background: #f8fafc; border-radius: 8px; margin-bottom: 15px;",
    h4("Adjust Debt Service and Interest Rate to see max loan", style = "margin-top: 0; color: #1e3a8a;"),
    fluidRow(
      column(6,
        sliderInput("ds_input", "Annual Debt Service Capacity",
                    min = 50000, max = 250000, value = 107000,
                    step = 5000, pre = "$", sep = ",",
                    width = "100%")
      ),
      column(6,
        sliderInput("rate_input", "Interest Rate (%)",
                    min = 4, max = 10, value = 7,
                    step = 0.25, post = "%",
                    width = "100%")
      )
    ),
    fluidRow(
      column(12, align = "center",
        p(style = "margin: 5px 0; color: #6b7280; font-size: 12px;",
          paste0("Current DS Capacity (80% of NOI @ 1.25x DSCR): ~", 
                 dollar(current_noi_ref * 0.8)))
      )
    )
  ),
  renderUI({
    ds <- input$ds_input
    rate <- input$rate_input
    
    # Calculate max loan
    max_loan <- ds / (rate / 100)
    
    # Color logic: Higher DS + Lower Rate = more borrowing power = GREEN
    # Lower DS + Higher Rate = less borrowing power = RED
    ds_score <- (ds - 50000) / (250000 - 50000)  # 0 to 1
    rate_score <- (10 - rate) / (10 - 4)  # Inverted: lower rate = higher score
    combined_score <- (ds_score + rate_score) / 2
    combined_score <- max(0, min(1, combined_score))
    
    # Color gradient
    if (combined_score < 0.5) {
      r <- 220
      g <- round(180 * (combined_score * 2))
      b <- 50
    } else {
      r <- round(220 - 180 * ((combined_score - 0.5) * 2))
      g <- 180
      b <- 50
    }
    bg_color <- sprintf("rgb(%d, %d, %d)", r, g, b)
    text_color <- if (combined_score > 0.3) "#1a1a1a" else "#ffffff"
    
    # Loan quality indicator
    loan_quality <- case_when(
      max_loan >= 2500000 ~ "Strong Capacity",
      max_loan >= 1800000 ~ "Good Capacity",
      max_loan >= 1200000 ~ "Moderate Capacity",
      max_loan >= 800000 ~ "Limited Capacity",
      TRUE ~ "Constrained"
    )
    
    div(style = "text-align: center; padding: 40px;",
      div(style = paste0(
        "display: inline-block; padding: 40px 60px; border-radius: 16px; ",
        "background: ", bg_color, "; ",
        "box-shadow: 0 10px 40px rgba(0,0,0,0.15);"
      ),
        p(style = paste0("margin: 0 0 10px 0; font-size: 16px; color: ", text_color, "; opacity: 0.8;"),
          "Maximum Loan Amount"),
        h1(style = paste0("margin: 0; font-size: 48px; font-weight: bold; color: ", text_color, ";"),
           dollar(max_loan)),
        p(style = paste0("margin: 10px 0 0 0; font-size: 14px; color: ", text_color, "; opacity: 0.9;"),
          loan_quality)
      ),
      div(style = "margin-top: 25px; color: #6b7280;",
        p(style = "margin: 5px 0;", 
          strong("Formula: "), "Max Loan = Debt Service / Interest Rate"),
        p(style = "margin: 5px 0; font-size: 13px;",
          HTML(paste0("Higher DS + Lower Rate = <span style='color: #16a34a; font-weight: bold;'>GREEN</span> (more capacity) | ",
                      "Lower DS + Higher Rate = <span style='color: #dc2626; font-weight: bold;'>RED</span> (constrained)")))
      )
    )
  })
)
```

Row {data-height=100}
-----------------------------------------------------------------------

### How to Use These Calculators

Use the sliders above to explore different scenarios:

**Portfolio Valuation**: Shows how NOI and Cap Rate combine to determine portfolio value. As you increase NOI and Cap Rate toward stabilization targets, the color shifts from red to green.

**Loan Affordability**: Shows maximum loan amount based on debt service capacity and interest rates. Higher debt service capacity and lower interest rates increase borrowing power (green).
