---
title: "Demographic & Housing Trend Intelligence"
subtitle: "Portfolio Performance Evaluation"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(leaflet)
library(DT)
library(scales)

final_data <- read_rds("../data/processed/final_analytical_data.rds")
properties <- final_data$property_reference
property_crosswalk <- final_data$property_crosswalk
tract_counts <- final_data$tract_counts
zcta_counts <- final_data$zcta_counts
county_counts <- final_data$county_counts
tract_history <- final_data$tract_history
zcta_history <- final_data$zcta_history
classifications <- final_data$tract_classifications
variable_dictionary <- final_data$variable_dictionary
spec_years <- range(tract_history$year, na.rm = TRUE)
```

## Executive Summary

This report analyzes demographic and housing trends for the areas containing the **`r nrow(properties)` properties**. It starts at the **county** level, steps down to **ZCTA**, and finishes with a concise **tract** view limited to tracts that actually contain our assets. ACS **5-year** data (2013–`r max(spec_years)`) and the latest **1-year** county data (through 2023) are inflation-adjusted to 2023 dollars; directional signals respect ACS margins of error (MOE) at the ~90% confidence level.

### Key Findings

```{r summary_stats}
prop_joined <- properties %>%
  left_join(classifications, by = c("tract_geoid" = "GEOID"))

summary_counts <- prop_joined %>%
  count(classification, name = "n_props") %>%
  mutate(pct = n_props / sum(n_props))

growth_pct <- summary_counts %>% filter(classification == "Growth") %>% pull(pct)
growth_pct <- ifelse(length(growth_pct) == 0, 0, growth_pct)
```

* **`r percent(growth_pct)`** of the portfolio is located in **Growth** tracts (significant population + real income gains with stable/improving vacancy).
* County metrics use the latest ACS 1-year data; tract/ZCTA use ACS 5-year series; all monetary values in **2023 dollars**.

## Spec Compliance Snapshot

- ACS **5-year** for tract/ZCTA (2013–`r max(spec_years)`); **1-year** county through 2023.
- MOEs propagated into derived metrics; significance tested at ~90% confidence.
- Custom age buckets (<18, 18–25, 26–55, 56–65, 65+); inflation-adjusted income/rent.

## Portfolio Map

```{r map}
pal <- colorFactor(c("green", "orange", "red"), domain = c("Growth", "Stable", "Weakening"))

leaflet(prop_joined) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~long, ~lat,
    color = ~ pal(classification),
    radius = 6,
    fillOpacity = 0.8,
    popup = ~ paste0(
      "<b>", full_address, "</b><br>",
      "Tract: ", tract_geoid, "<br>",
      "Class: ", classification, "<br>",
      "Pop Change: ", percent(pop_change, 0.1), "<br>",
      "Real Income Change: ", percent(income_change_real, 0.1)
    )
  ) %>%
  addLegend("bottomright", pal = pal, values = ~classification, title = "Neighborhood Trend")
```

## Property–Census Crosswalk & Counts

```{r crosswalk}
datatable(property_crosswalk, options = list(pageLength = 10), caption = "Property-to-Census crosswalk (tract, ZCTA, county)")
```

```{r crosswalk_counts}
datatable(tract_counts, options = list(pageLength = 10), caption = "Property count by Census tract")
datatable(zcta_counts, options = list(pageLength = 10), caption = "Property count by ZCTA")
datatable(county_counts, options = list(pageLength = 10), caption = "Property count by county")
```

## County View (Jefferson County, AL)

```{r county_prep}
county_panel <- final_data$county_history_1yr
if (nrow(county_panel) == 0) county_panel <- final_data$county_history_5yr
county_year_breaks <- scales::pretty_breaks()(range(county_panel$year))
```

```{r county_plots}
county_panel %>%
  ggplot(aes(x = year, y = total_popE)) +
  geom_line(color = "#1b9e77", size = 1.1) +
  scale_x_continuous(breaks = county_year_breaks, labels = scales::number_format(accuracy = 1)) +
  theme_minimal() +
  labs(title = "County Population (ACS, latest available)", y = "Population", x = "Year")

county_panel %>%
  ggplot(aes(x = year, y = med_income_real)) +
  geom_line(color = "#7570b3", size = 1.1) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = county_year_breaks, labels = scales::number_format(accuracy = 1)) +
  theme_minimal() +
  labs(title = "County Real Median Household Income (2023$)", y = "Income", x = "Year")

county_panel %>%
  ggplot(aes(x = year, y = vacancy_rate)) +
  geom_line(color = "#e7298a", size = 1.1) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = county_year_breaks, labels = scales::number_format(accuracy = 1)) +
  theme_minimal() +
  labs(title = "County Housing Vacancy Rate", y = "Vacancy Rate", x = "Year")
```

## ZCTA (Zip) View

```{r zcta_prep}
relevant_zips <- properties$zcta5 %>% coalesce(properties$zip) %>% unique() %>% na.omit()
zcta_panel <- zcta_history %>% filter(GEOID %in% relevant_zips)
latest_zcta <- zcta_panel %>% group_by(GEOID) %>% filter(year == max(year)) %>%
  select(GEOID, med_income_real, vacancy_rate, pct_bachelors, poverty_rate, med_rent_real)
```

```{r zcta_plots}
if (nrow(zcta_panel) > 0) {
  zcta_breaks <- scales::pretty_breaks()(range(zcta_panel$year))

  zcta_panel %>%
    ggplot(aes(x = year, y = med_income_real, group = GEOID, color = GEOID)) +
    geom_line(alpha = 0.6, size = 1) +
    scale_y_continuous(labels = dollar) +
    scale_x_continuous(breaks = zcta_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    guides(color = "none") +
    labs(title = "Zip-Level Real Median Income (2023$)", y = "Income", x = "Year")

  zcta_panel %>%
    ggplot(aes(x = year, y = vacancy_rate, group = GEOID, color = GEOID)) +
    geom_line(alpha = 0.6, size = 1) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = zcta_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    guides(color = "none") +
    labs(title = "Zip-Level Vacancy Rate", y = "Vacancy Rate", x = "Year")
}
```

```{r zcta_table}
if (nrow(zcta_panel) > 0) {
  latest_zcta %>%
    mutate(
      med_income_real = dollar(med_income_real),
      med_rent_real = dollar(med_rent_real),
      vacancy_rate = percent(vacancy_rate, 0.1),
      pct_bachelors = percent(pct_bachelors, 0.1),
      poverty_rate = percent(poverty_rate, 0.1)
    ) %>%
    datatable(options = list(pageLength = 10), caption = "Latest available ZCTA metrics")
}
```

## Tract View (Concise)

```{r tract_prep}
relevant_tracts <- unique(prop_joined$tract_geoid)
tract_panel <- tract_history %>% filter(GEOID %in% relevant_tracts)
tract_year_breaks <- scales::pretty_breaks()(range(tract_panel$year))
tract_summary <- final_data$tract_classifications %>%
  filter(GEOID %in% relevant_tracts) %>%
  left_join(tract_panel %>% group_by(GEOID) %>% summarise(latest_year = max(year)), by = "GEOID") %>%
  arrange(classification, desc(pop_cagr))

spotlight <- tract_summary %>%
  arrange(desc(pop_cagr)) %>%
  slice_head(n = 3)
spotlight_panel <- tract_panel %>% filter(GEOID %in% spotlight$GEOID)
```

### Tract Summary Table

```{r tract_table}
tract_summary %>%
  transmute(
    Tract = GEOID,
    Classification = classification,
    `Pop CAGR` = percent(pop_cagr, 0.1),
    `Income CAGR (real)` = percent(income_cagr, 0.1),
    `Vacancy Δ` = percent(vacancy_change, 0.1),
    `Pop Change Sig` = ifelse(pop_change_sig, "Yes", "No"),
    `Income Change Sig` = ifelse(income_change_sig, "Yes", "No"),
    `Vacancy Change Sig` = ifelse(vacancy_change_sig, "Yes", "No")
  ) %>%
  datatable(options = list(pageLength = 10), caption = "Tract-level classification and trend significance")
```

### Spotlight Tracts (top movers)

```{r tract_spotlight_plots}
if (nrow(spotlight_panel) > 0) {
  spotlight_panel %>%
    ggplot(aes(x = year, y = total_popE, color = GEOID)) +
    geom_line(size = 1.1) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Population (Spotlight Tracts)", y = "Population", x = "Year", color = "Tract")

  spotlight_panel %>%
    ggplot(aes(x = year, y = med_income_real, color = GEOID)) +
    geom_line(size = 1.1) +
    scale_y_continuous(labels = dollar) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Real Median Income (Spotlight Tracts)", y = "Income (2023$)", x = "Year", color = "Tract")
}
```

## Age Composition (Tract-Level, Portfolio Tracts Only)

```{r age_comp}
if (nrow(tract_panel) > 0) {
  age_agg <- tract_panel %>%
    group_by(year) %>%
    summarise(
      pct_under18 = mean(pct_age_under18, na.rm = TRUE),
      pct_18_25 = mean(pct_age_18_25, na.rm = TRUE),
      pct_26_55 = mean(pct_age_26_55, na.rm = TRUE),
      pct_56_65 = mean(pct_age_56_65, na.rm = TRUE),
      pct_65plus = mean(pct_age_65plus, na.rm = TRUE)
    ) %>%
    pivot_longer(-year, names_to = "bucket", values_to = "value")

  ggplot(age_agg, aes(x = year, y = value, color = bucket, group = bucket)) +
    geom_line(size = 1.1) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Average Age Composition (Portfolio Tracts)", y = "Percent", x = "Year", color = "Age Bucket")
}
```

## Education & Economic Status (Tract-Level, Portfolio Tracts Only)

```{r edu_econ}
if (nrow(tract_panel) > 0) {
  edu_pov <- tract_panel %>%
    group_by(year) %>%
    summarise(
      pct_bachelors = mean(pct_bachelors, na.rm = TRUE),
      pct_hs_or_higher = mean(pct_hs_or_higher, na.rm = TRUE),
      poverty_rate = mean(poverty_rate, na.rm = TRUE),
      unemployment_rate = mean(unemployment_rate, na.rm = TRUE),
      med_income_real = mean(med_income_real, na.rm = TRUE)
    )

  edu_long <- edu_pov %>%
    select(year, pct_hs_or_higher, pct_bachelors) %>%
    pivot_longer(-year, names_to = "metric", values_to = "value")

  ggplot(edu_long, aes(x = year, y = value, color = metric)) +
    geom_line(size = 1.1) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Education Attainment (Portfolio Tracts)", y = "Percent", x = "Year", color = "Metric")

  econ_long <- edu_pov %>%
    select(year, poverty_rate, unemployment_rate) %>%
    pivot_longer(-year, names_to = "metric", values_to = "value")

  ggplot(econ_long, aes(x = year, y = value, color = metric)) +
    geom_line(size = 1.1) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Poverty & Unemployment (Portfolio Tracts)", y = "Percent", x = "Year", color = "Metric")

  ggplot(edu_pov, aes(x = year, y = med_income_real)) +
    geom_line(color = "#1f78b4", size = 1.1) +
    scale_y_continuous(labels = dollar) +
    scale_x_continuous(breaks = tract_year_breaks, labels = scales::number_format(accuracy = 1)) +
    theme_minimal() +
    labs(title = "Average Real Median Household Income (Portfolio Tracts)", y = "Income (2023$)", x = "Year")
}
```

## Property Performance vs. Market

```{r performance_table}
prop_joined %>%
  select(Address = address_raw, Class = classification, `Acq Cost` = acquisition_cost, `FMV` = fair_market_value) %>%
  datatable()
```

## Methodology

* **Data Source**: US Census Bureau, ACS 5-year (tract/ZCTA) 2013–`r max(spec_years)`; ACS 1-year county through 2023.
* **Inflation Adjustment**: All monetary values adjusted to 2023 dollars using CPI-U annual averages.
* **MOE Handling**: ACS MOEs propagated through ratios (SE = MOE/1.645); trend direction requires differences to exceed combined MOE.
* **Classification**:
  * **Growth**: Statistically significant population increase and ≥1% real income CAGR with stable or declining vacancy.
  * **Weakening**: Significant population or income decline, or significant vacancy increase (>1 pp).
  * **Stable**: No statistically significant movement on core indicators.
* **Custom Age Buckets**: Allocation from ACS 5-year age bands to <18, 18–25, 26–55, 56–65, 65+.
* **Variable Catalog**:

```{r var_table, results='asis'}
if (nrow(variable_dictionary) > 0) {
  variable_dictionary %>%
    knitr::kable()
}
```
